mathml: Translate R expressions to MathML and LaTeX/MathJax

### 2022-12-19 Draft3.Rmd
R\ : added backslash after R

The R extension of the markdown language [@Xie2020] enables reproducible
statistical reports with nice typesetting in HTML, Microsoft Word, and Latex.
Since recently [@R, version 4.2], R's manual pages include support for
mathematical expressions [@Sarkar2022], which already is a big improvement.
However, rules for the translation of R's built-in language elements to their
mathematical representation are still lacking. So far, R expressions such
as `pbinom(k, N, p)` are printed as they are; pretty mathematical formulae
like $P_\mathrm{Bi}(X \le k; N, p)$ require explicit Latex commands, that
is, `P_\mathrm{Bi}\left(X \le k; N, p\right)`. Except for minimalistic examples,
these commands are tedious to type in and hard to read in their source code.

The present R package defines a set of rules for an _automatic_ translation of
R expressions to mathematical output in RMarkdown documents [@Xie2020] and
ShinyApp webpages [@Chang2022]. The translation is done by an embedded Prolog
interpreter that maps nested expressions recursively to MathML and
LaTeX/MathJax, respectively. User-defined hooks enable to extend the set of
rules, for example, to represent specific R elements by custom mathematical
signs.

The main feature of the package that the same R expressions and equations are
used for both mathematical typesetting and calculations. This saves time and
reduces mistakes, as well be illustrated below.

The paper is organized as follows. Section\ 2 presents the background of the
package, including Prolog and the R package \pkg{rolog} and the two main classes
of rules for translating R\ objects to mathematical expressions. Section\ 3
illustrates the main features of the \pkg{mathml} package, potential issues and
workarounds using examples from the day-to-day perspective of the user.
Sections\ 4 and\ 5 present two case studies for the use of the package in a
scientific report and a ShinyApp web page. Section\ 6 concludes with a
discussion and ideas for further development. A complete list of the R elements
covered by the package is found in the Appendix.

@@

The translation of R expressions to mathematical output is achieved through a
Prolog interpreter. Prolog is a classical logic programming language with many
applications in expert systems, computer linguistics and symbolic artificial
intelligence. The main strength of Prolog is its concise representation of facts
and rules for the representation of knowledge and grammar, as well as its
efficient built-in search engine for closed world domains. As it is well-known,
R\ is a statistical programming language for data analysis and statistical
modeling which is widely used in academia and industry. Besides the core
library, a lot of packages have been developed for all kinds of statistical
problems, including statistics-based artificial intelligence tools such as
neural networks for machine learning and deep learning. Whereas Prolog is weak
in statistical computation, but strong in symbolic manipulation, the converse
may be said for the R language. The \pkg{rolog} package bridges this gap by
providing an interface to a SWI-Prolog distribution in R. The communication
between the two systems is mainly in the form of queries from R\ to Prolog, but
two predicates allow Prolog to ring back and evaluate terms in R.

@@

The R package \pkg{mathml} can now be used to render the call in MathML, that
is the math dialect for HTML webpages or in MathJax/LaTeX, as shown
below (some of the curly braces are not really needed in this simple example,
but are necessary in edge cases).

@@

We can include the output in a RMarkdown document by specifying `results='asis'`
in the R code chunk, as is done in the next example. The
R\ function `mathout(...)` is just a wrapper that invokes `mathml(...)` for HTML
output and `mathjax(...)` for LaTeX output.

@@

As can be seen, \pkg{mathml} reminds the user that it is not possible to change
the precedence of custom operators in R, and that expressions with such
operators are evaluated strictly from left to right. As in the
previous example, the solution is to either use the functional form of the
operator or a curly brace to enforce the correct operator precedence.

@@

R's `integrate` function takes a number of arguments, the most important ones
being the function to integrate, and the lower and the upper bound of the
integration. For mathematical typesetting in the form
of $\int f(x)\, dx$, \pkg{mathml} needs to find out the name of the integration
variable. For that purpose, the underlying Prolog bridge provides a
predicate `r_eval/2` that calls R from Prolog. In the example below, this
predicate is used for a call of `formalArgs(args(sin))`, which returns the names
of the function arguments of `sin`, namely, `x`.

@@

The extra R function `canonical()` applies `match.call()` to non-primitive R
calls, basically cleaning up the arguments and bringing them into the correct
order.

@@

An interesting use case of the package is in this sense the readability of self-written
functions in R code. R code is notoriously hard to read, as are most programming
languages [@green1977]. The poor legibility of the language is one of the main
sources of mistakes. For illustration, we consider Equation 10 in
Schwarz [-@schwarz1994], namely the same equation reported above. The first version has a wrong parenthesis, which is barely visible in the R code.

>>>>

The R\ extension of the markdown language [@Xie2020] enables reproducible
statistical reports with nice typesetting in HTML, Microsoft Word, and Latex.
Since recently [@R, version 4.2], R\'s manual pages include support for
mathematical expressions [@Sarkar2022], which already is a big improvement.
However, rules for the translation of R\'s built-in language elements to their
mathematical representation are still lacking. So far, R\ expressions such
as `pbinom(k, N, p)` are printed as they are; pretty mathematical formulae
like $P_\mathrm{Bi}(X \le k; N, p)$ require explicit Latex commands, that
is, `P_\mathrm{Bi}\left(X \le k; N, p\right)`. Except for minimalistic examples,
these commands are tedious to type in and hard to read in their source code.

The present R\ package defines a set of rules for an _automatic_ translation of
R\ expressions to mathematical output in RMarkdown documents [@Xie2020] and
ShinyApp webpages [@Chang2022]. The translation is done by an embedded Prolog
interpreter that maps nested expressions recursively to MathML and
LaTeX/MathJax, respectively. User-defined hooks enable to extend the set of
rules, for example, to represent specific R\ elements by custom mathematical
signs.

The main feature of the package that the same R\ expressions and equations are
used for both mathematical typesetting and calculations. This saves time and
reduces mistakes, as well be illustrated below.

The paper is organized as follows. Section\ 2 presents the background of the
package, including Prolog and the R\ package \pkg{rolog} and the two main classes
of rules for translating R\ objects to mathematical expressions. Section\ 3
illustrates the main features of the \pkg{mathml} package, potential issues and
workarounds using examples from the day-to-day perspective of the user.
Sections\ 4 and\ 5 present two case studies for the use of the package in a
scientific report and a ShinyApp web page. Section\ 6 concludes with a
discussion and ideas for further development. A complete list of the R\ elements
covered by the package is found in the Appendix.

@@

The translation of R\ expressions to mathematical output is achieved through a
Prolog interpreter. Prolog is a classical logic programming language with many
applications in expert systems, computer linguistics and symbolic artificial
intelligence. The main strength of Prolog is its concise representation of facts
and rules for the representation of knowledge and grammar, as well as its
efficient built-in search engine for closed world domains. As it is well-known,
R\ is a statistical programming language for data analysis and statistical
modeling which is widely used in academia and industry. Besides the core
library, a lot of packages have been developed for all kinds of statistical
problems, including statistics-based artificial intelligence tools such as
neural networks for machine learning and deep learning. Whereas Prolog is weak
in statistical computation, but strong in symbolic manipulation, the converse
may be said for the R\ language. The \pkg{rolog} package bridges this gap by
providing an interface to a SWI-Prolog distribution in R. The communication
between the two systems is mainly in the form of queries from R\ to Prolog, but
two predicates allow Prolog to ring back and evaluate terms in R.

@@

The R\ package \pkg{mathml} can now be used to render the call in MathML, that
is the math dialect for HTML webpages or in MathJax/LaTeX, as shown
below (some of the curly braces are not really needed in this simple example,
but are necessary in edge cases).

@@

We can include the output in a RMarkdown document by specifying `results='asis'`
in the R\code chunk, as is done in the next example. The
R\ function `mathout(...)` is just a wrapper that invokes `mathml(...)` for HTML
output and `mathjax(...)` for LaTeX output.

@@

As can be seen, \pkg{mathml} reminds the user that it is not possible to change
the precedence of custom operators in R\, and that expressions with such
operators are evaluated strictly from left to right. As in the
previous example, the solution is to either use the functional form of the
operator or a curly brace to enforce the correct operator precedence.

@@

R\'s `integrate` function takes a number of arguments, the most important ones
being the function to integrate, and the lower and the upper bound of the
integration. For mathematical typesetting in the form
of $\int f(x)\, dx$, \pkg{mathml} needs to find out the name of the integration
variable. For that purpose, the underlying Prolog bridge provides a
predicate `r_eval/2` that calls R\ from Prolog. In the example below, this
predicate is used for a call of `formalArgs(args(sin))`, which returns the names
of the function arguments of `sin`, namely, `x`.

@@

The extra R\ function `canonical()` applies `match.call()` to non-primitive R\
calls, basically cleaning up the arguments and bringing them into the correct
order.

@@

An interesting use case of the package is in this sense the readability of self-written
functions in R\ code. R\ code is notoriously hard to read, as are most programming
languages [@green1977]. The poor legibility of the language is one of the main
sources of mistakes. For illustration, we consider Equation 10 in
Schwarz [-@schwarz1994], namely the same equation reported above. The first version has a wrong parenthesis, which is barely visible in the R code.
